|

### 2022-09-04 submission/reviewer_response.rmd

---
title: "Response to reviewers"
output:
  bookdown::pdf_document2: default
geometry: margin=1.0in
toc: true
font-size: 12pt
editor_options: 
  chunk_output_type: console
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
bibliography: references.bib
csl: jce.csl
---

# Reviewer 1 {-}
"Individualized treatment effect was predicted best by modeling baseline risk in
interaction with treatment assignment"
Comments to the Authors:

I read the manuscript with interest, however, the authors first need to address
the following concerns.

- Page 2.  Line 6. Need to clarify what it means by baseline risk, as well as
what it means by the “reference.”  Line 10. Related to the comment above, can
you provide a sentence or two briefly describing what risk-modeling approaches
are?  Line 12. Need to clarify what it means by “absolute benefit.”
Line 13. Not clear, from the written sentence, which methods (whether they are
risk-based methods or the method based on modeling treatment effect-modifiers)
lead to the weaker discrimination of benefit in the presence of true effect
modifiers.  Line 18. The estimates refer to estimates of which parameters? Need
to clarify. Also, please indicate why these estimates may not be applied to
individual patients. \textcolor{green}{[DONE]}

- Page 4.  Line 1. Not clear what it means by the "AUC value of 0.75." Need to
explain AUC. And after all, what is the "true" value of $\beta$ that was used in the
simulations?  Line 6. What is c? And which term in the equation is the term that
corresponds to the constant treatment effect? What are the values of vs used in
the simulations?  Line 11. What do you mean by “standard” alpha? what is
"standard"? What is "effect"? Does it refer to the effect on the approaches'
performance? What do you mean by the effect of risk model discriminative
ability? How can risk model discriminative ability have any effects? Please
clarify. \textcolor{green}{[DONE]}

- Page 5.  Line 14. Please clarify how $\delta_0$, $\delta_1$ and $\delta_3$ are
  estimated. \textcolor{green}{[DONE]}

- Page 6.  Line 8. Please describe in a sentence what "ICI for benefit" is.
Line17. Please comment on how those 6 predictors were selected (is it also
selected based on AIC?). \textcolor{green}{[DONE]}

-  Page 7.  Line 14. Please describe how the "absolute benefit predictions" were
evaluated (please present the evaluation metric). \textcolor{green}{[DONE]}

- Page 8. In the figures, I think for each panel you should label the X-axis
(i.e., the methods under consideration). At first, it was difficult to
understand what was displayed. \textcolor{green}{[DONE]}

- Page 13. In Figure 4, again, what is the C-statistic for benefit? Need an
explicit expression or a clear description (both in figure caption and in the
text). \textcolor{green}{[DONE]}

- Page 16. In Figure 6, please also display the model-based confidence bands for
all the methods. Also please clarify how the predicted benefit was
computed. Does it just mean the treatment effect? If so, how could the "constant
treatment effect" method have a non-constant treatment effect as a function of
the baseline risk? Please clarify. \textcolor{green}{[DONE]}

- Page 18.  Line 21. Generally, in the estimation, how do you distinguish the
treatment-related harm from the actual treatment effect? \textcolor{red}{[UNRESOLVED]}

- Page 19.  Line 1. As discussed in the Discussion, the authors only considered
the risk-based methods. Along this line, I think the title of the manuscript is,
to some extent, misleading, because it gives an impression that the modeling
baseline risk-and-treatment interaction is a “globally best method” for the
individualized treatment effect estimation, which is not likely to be true.
Line 2. Why only tree-based approaches were cited? There is a large volume of
treatment effect modeling methods, which I think should be cited as well, if
tree-based approaches are to be cited.  Line 8. I don’t think what is stated in
line 8 is true when the "true" data model is defined based on separate
treatment-by-covariate interaction terms, in addition to the main effect
terms. Furthermore, there are many methods that avoid overfitting through
regularization and simultaneous treatment effect modifier selection. Authors
should discuss this. \textcolor{red}{[UNRESOLVED]}

# Reviewer 2 {-}

I reviewed the same manuscript last month. It seems that the manuscript is not
revised and is the same as the previous version and did not address any of the
comments. Please see my comments made in the previous round of review. Would you
please check if this attached manuscript is a revised version ? Thank you.

# Reviewer 3 {-}

The content of this paper is methodologically sound and the results
are consistent with what the literature has shown (namely that more complex
models are often outperformed by simpler ones in small sample and weaker signal
scenarios) and the writing is generally adequate. However, as written, the paper
does not focus on the significance of this methodology and the simulation
studies in terms of guiding clinical care (see Comment 1 for more details); this
was the main weakness of this paper.

1. The introduction will benefit from more emphasis on the practical
   significance of modeling baseline risk and treatment effects. In particular,
   in which real-life scenarios was this type of analysis useful and what
   happened when the baseline risk and its interaction with treatment was
   neglected in favor of a simpler approach? What type of analyses are typically
   done in such scenarios and what does the approach presented in this
   manuscript offer on top of the usual analysis methods?

2. During the application of restricted cubic splines, was there an assumption
   of interaction between risk and treatment effects as well? For example, were
   the restricted cubic splines fit separately to each treatment arm?

3. It would also be helpful to discuss the results (particularly the fact that
   the simpler model--linear interaction--outperforms RCS-3 in smaller sample
   size scenarios and weaker signals, even when there is nonlinearity in the
   true underlying relationship, as well as the fact that RCS-3 outperforms
   RCS-4 and RCS-5 in many scenarios) in terms of the bias-variance tradeoff. In
   particular, the restricted cubic splines-based methods introduce additional
   degrees of freedom and thus increase the variance, despite being expected to
   reduce the bias.

# Reviewer 4 {-}

This paper presents the results from a very large simulation study to
evaluate a number of fairly simple statistical approaches to estimate individual
treatment effects using data from a RCT. The paper is well written and the
results are presented objectively. The results show essentially what you might
expect, that if the data generating method is more complicated then flexible
methods perform best, while if the data generation method is more simple then
simpler methods do best. Furthermore, with larger sample size flexible methods
tend to do better. The authors end up essentially recommending an approach that
has a linear interaction term, or one with a cubic spline. A couple of the
comparators are even more simple than these (constant treatment effect and
stratification by quartiles), and it is no surprise that these naïve method are
not good in the simulations. So if the point of the paper is to discourage use
of these naïve methods, then that is a valuable contribution.  The big
limitation of the whole study is that the baseline risk (i.e. the linear
combination of the X variables from the control arm of the RCT) is assumed to be
the driver of the individual treatment effect. The data is generated with this
assumption, and all the estimation methods make this assumption. The authors
totally acknowledge this limitation. But I would have thought that in the
majority of realistic situations the linear combination of X’s for the main
effect would have been different from the linear combination of X’s in the
interaction with treatment. In this case the authors conclusion that linear
interaction and RCS3 are good, would likely not hold up. The authors do
reference a few articles that use tree-based methods (there are many many more
and lots of non tree-based approaches too), and correctly point out their
potential for overfitting. But regularization and cross-validation approaches
can do a pretty good job at limiting the overfitting. So I think I would favor
tree-based (or other more modern approaches), unless there is some clinical or
scientific rationale to believe that the control group risk is the driver of the
treatment effect heterogeneity.  A couple of others things that should be
mentioned as caveats. In the simulated data the X variables are made independent
of each other. If they were made to be correlated, that would probably be more
realistic. Do you have any information on how this might change the findings, in
general terms.  Larger sample sizes also opens up the possibility of
generalizing the linear additive assumption for the control group risk. For
example, this part of the model could now include non-linear terms in X and
interactions between X’s.

### 2022-09-04 submission/manuscript_tracked.rmd
drafting changes based on reviewer comments

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
author:
  - Alexandros Rekkas$^1$,
  - Peter R. Rijnbeek$^1$,
  - David M. Kent$^2$,
  - Ewout W. Steyerberg$^3$,
  - David van Klaveren$^4$
numbersections: true
linenumbers: false
output:
  bookdown::pdf_document2: default
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

# Introduction
Predictive approaches to heterogeneity of treatment effects (HTE) aim at the
development of models predicting either individualized effects or which of two
(or more) treatments is better for an individual [@Varadhan2013]. In prior work,
we divided such methods in three broader categories based on the reference class
used for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
baseline risk as the reference; treatment effect modeling approaches also model
treatment-covariate interactions, in addition to risk factors; optimal treatment
regime approaches focus on developing treatment assignment rules and rely
heavily on modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling treatment-covariate interactions, often led
to miscalibrated predictions of absolute benefit, contrary to risk-based
methods, despite their weaker discrimination of benefit in the presence of true
effect modifiers [@vanKlaveren2019]. Most often, risk-modeling approaches are
carried out in two steps: first a risk prediction model is developed externally
or internally on the entire RCT population, “blinded” to treatment; then the RCT
population is stratified using this prediction model to evaluate risk-based
treatment effect variation [@Kent2010]. This approach identified substantial
absolute treatment effect differences between low-risk and high-risk patients in
a re-analysis of 32 large trials [@Kent2016]. However, even though estimates at
the risk subgroup level may be accurate, these estimates may not apply to
individual patients.

In the current simulation study, we aim to summarize and compare different
risk-based models for predicting treatment effects. We simulate different
relations between baseline risk and treatment effects and also consider
potential harms of treatment. We illustrate the different models by a case study
of predicting individualized effects of treatment for acute myocardial
infarction (MI) in a large RCT.

@@

We simulated a typical RCT, comparing equally-sized treatment and control arms
in terms of a binary outcome. For each patient we generated 8 baseline
covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim
B(1,0.2)$. Outcomes in the control arm were generated from Bernoulli variables
with true probabilities following a logistic regression model including all
baseline covariates, i.e.
$P(Y(0)=1\given X=x) = \text{expit}(lp_0) = e^{lp_0}/(1+e^{lp_0})$,
with $lp_0=lp_0(x)=x^t\beta$.
In the base scenarios coefficient values $\beta$ were
such, that the AUC of the logistic regression model was 0.75 and the event rate
in the control arm was $20\%$.

Outcomes in the treatment arm were first generated using 3 simple scenarios:
absent (OR = 1), moderate (OR = 0.8) or strong (OR = 0.5) constant relative
treatment effect. We then introduced linear, quadratic and non-monotonic
deviations from constant treatment effects using:
$$lp_1 = \gamma_2(lp_0-c)^2 + \gamma_1(lp_0-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(Y(1)=1\given X=x) = \text{expit}(lp_1)$.
Finally, we incorporated constant absolute harms for all treated patients, such
that
$P(Y(1)=1|X=x) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 (80\% power for the
detection of a marginal OR of 0.8 with the standard alpha of 5\%). We evaluated
the effect of smaller or larger sample sizes of 1,063 and 17,000,
respectively. We also evaluated the effect of risk model discriminative ability,
adjusting the baseline covariate coefficients, such that the AUC of the
regression model in the control arm was 0.65 and 0.85, respectively.

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant odds ratio). Hence, absolute benefit is predicted from
$\tau(x;\hat{\beta}) = \text{expit}(\hat{lp}_0) - \text{expit}(\hat{lp}_0+\delta_0)$, where
$\delta_0$ is the log of the assumed constant odds ratio and
$\hat{lp}_0 = \hat{lp}_0(x;\hat{\beta}) = x^t\hat{\beta}$
the linear predictor of the estimated baseline risk model.

Third, we considered a logistic regression model including treatment, the
prognostic index, and their linear interaction. Absolute benefit is then
estimated from
$\tau(x;\hat{\beta}) = \text{expit}(\delta_0+\delta_1\hat{lp}_0) - \text{expit}(\delta_0+\delta_2+(\delta_1+\delta_3)\hat{lp}_0)$
We will refer to this method as the *linear interaction* approach.

@@

We compared the discriminative ability of the methods under study using
c-for-benefit and the integrated calibration index (ICI) for
benefit (Supplement, Section 6) [@vanKlaveren2018].

@@

We demonstrated the different methods using `r nrow(gusto)` patients with acute
myocardial infarction (MI) included in the GUSTO-I trial.  `r treatmentArms %>%
filter(treatment == 1) %>% pull(n)` patients were randomized to tissue
plasminogen activator (tPA) treatment and `r treatmentArms %>% filter(treatment
== 0) %>% pull(n)` were randomized to streptokinase. The outcome of interest
was 30-day mortality (total of `r sum(gusto$outcome)` events), recorded for all
patients. In line with previous analyses [@Califf1997; @Steyerberg2000], we
fitted a logistic regression model with 6 baseline covariates, i.e. age, Killip
class, systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk (Supplement, Section 8).

@@

The constant treatment effect approach outperformed other approaches in the base
case scenario (N = `r 4250`; OR = 0.8; AUC= 0.75; no absolute treatment harm)
with a true constant treatment effect (median RMSE: constant treatment effect 
`r median(rmseBaseCase$constant_treatment_effect)`;
linear interaction `r median(rmseBaseCase$linear_predictor)`;
RCS-3 `r median(rmseBaseCase$rcs_3_knots)`). 
The linear interaction model was optimal under true linear deviations 
(median RMSE: constant treatment effect 
`r median(rmseLinearInteraction$constant_treatment_effect)`;
linear interaction `r median(rmseLinearInteraction$linear_predictor)`;
RCS-3 `r median(rmseLinearInteraction$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
and even in the presence of true quadratic deviations
(median RMSE: constant treatment effect 
`r median(rmseQuadratic$constant_treatment_effect)`;
linear interaction `r median(rmseQuadratic$linear_predictor)`;
RCS-3 `r median(rmseQuadratic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
from a constant relative treatment effect. With non-monotonic deviations, RCS-3
slightly outperformed the linear interaction model
(median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
With strong treatment-related harms the results were very similar in most
scenarios (Figure \ref{fig:rmsebase} panels A-C). Under non-monotonic deviations
the optimal performance of RCS-3 was more pronounced
(median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
A stronger average treatment effect (OR=0.5) led to larger absolute benefit
predictions and consequently to larger RMSE for all approaches, but the relative
differences between different approaches were similar to the base case scenario
(Supplement, Figure S10).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot on the top. The 2.5, 25, 50, 75, and 97.5 percentiles of the true benefit distributions are expressed by the boxplots on the side of the right-handside panel.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

@@

```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250. True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

All considered methods provided similar fits, predicting increasing benefits for
patients with higher baseline risk predictions, and followed the evolution of
the stratified estimates closely (Figure \ref{fig:gusto}). The constant
treatment effect model had somewhat lower AIC compared to the linear interaction
model (AIC: 
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(aic) %>% round()` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(aic) %>% round()`),
equal cross-validated discrimination
(c-for-benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(discrimination)`),
and slightly better cross-validated calibration
(ICI-for benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(calibration)` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(calibration)`).
In conclusion, although the sample size (`r nrow(gusto)` patients; 
`r sum(gusto$outcome)` events) allowed for flexible modeling approaches, a 
simpler constant treatment effect model is adequate for predicting absolute
30-day mortality benefits of treatment with tPA in patients with acute MI.

>>>>

---
title: |
 Individualized treatment effect was predicted best by modelin
 baseline risk in interaction with treatment assignment
abstract: |
author:
  - Alexandros Rekkas$^1$,
  - Peter R. Rijnbeek$^1$,
  - David M. Kent$^2$,
  - Ewout W. Steyerberg$^3$,
  - David van Klaveren$^4$
numbersections: true
linenumbers: false
output:
  bookdown::pdf_document2: default
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

# Introduction
Predictive approaches to heterogeneity of treatment effects (HTE) aim at the
development of models predicting either individualized effects or which of two
(or more) treatments is better for an individual

:::{custom-style="bluefont"}
with regard to a specific outcome of interest
:::

[@Varadhan2013]. In prior work, we divided such
methods in three broader categories based on the reference
class

:::{custom-style="bluefont"}
, i.e. the set of covariates} used for defining patient
:::

similarity when making individualized predictions or recommendations
[@Rekkas2020]. ~~Risk-modeling approaches use prediction of baseline risk as the
reference;~~

:::{custom-style="bluefont"}
Risk modeling methods predict similar treatment benefits for patients with
similar baseline outcome risk, i.e. the probability of a certain patient
experiencing the outcome of interest under the control;
:::

treatment effect modeling approaches also model treatment-covariate
interactions, in addition to risk factors; optimal treatment regime approaches
focus on developing treatment assignment rules and rely heavily on modeling
treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE].

:::{custom-style="bluefont"}
Risk modeling approaches are not new [@Rekkas2020] and are quite intuitive to
practitioners. Often medical guidelines rely on a risk stratified approach to
target treatments to different patients. In addition, re-analyses of studies
that only looked at overall results using risk stratification often result to
important insight on how treatment effects varied for diffrent patients. For
example, a risk stratified analysis of patients with acute myocardial infarction
(MI) based on the Thrombolysis in Myocardial Infarction (TIMI) risk score found
no benefit for patients who underwent primary angioplasty compared to
fibrynolysis. However, there was a significant benefit for patients with a high
TIMI score [@Thune2005]. Infants at lower risk of bronchopulmonary dysplasia
benefit relatively more from vitamin A therapy than infants at higher risk
[@Rysavy2021]. Finally, higher risk prediabetic patients benefit relatively more
from metformin than lower risk patients [@Sussman2015].
:::

:::{custom-style="bluefont"}
They result in increased power for detecting HTE, as they
evaluate interactions of treatment with a summary score, i.e. baseline outcome
risk, and, thus, avoid modeling treatment-covariate interactions.
:::

In simulations, modeling treatment-covariate interactions, often led to
miscalibrated predictions of ~~absolute benefit~~

:::{custom-style="bluefont"}
benefit on
the absolute scale (risk difference), ~~contrary~~ compared
:::

to risk-modeling methods~~, despite their weaker discrimination of benefit in
the presence of true effect modifiers~~
[@vanKlaveren2019].

:::{custom-style="bluefont"}
However, in the presence of true
treatment-covariate interactions, risk modeling methods performed worse in
separating lower from higher benefit patients compared to treatment effect
modeling methods.
:::

Most often, risk-modeling approaches are carried out in two
steps: first a risk prediction model is developed externally or internally on
the entire RCT population, “blinded” to treatment; then the RCT population is
stratified using this prediction model to evaluate risk-based treatment effect
variation [@Kent2010]. This approach identified substantial absolute treatment
effect differences between low-risk and high-risk patients in a re-analysis of
32 large trials [@Kent2016]. However, even though

:::{custom-style="bluefont"}
treatment effect estimates at the risk subgroup level may be accurate, these
estimates may not apply to individual patients,{as homogeneity of treatment
effects is assumed within risk strata. With stronger treatment effect
heterogeneity (larger variability between estimated risk-subgroup treatment
effects), predicted treatment effects for patients with similar predicted risk
may be quite different, if they end up in different risk strata.
:::

In the current simulation study, we aim to summarize and compare different
risk-based models for predicting treatment effects. We simulate different
relations between baseline risk and treatment effects and also consider
potential harms of treatment. We illustrate the different models by a case study
of predicting individualized effects of treatment for acute myocardial
infarction ~~(MI)~~in a large RCT.

@@

We simulated a typical RCT, comparing equally-sized treatment and control arms
in terms of a binary outcome. For each patient we generated 8 baseline
covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim
B(1,0.2)$. Outcomes in the control arm were generated from Bernoulli variables
with true probabilities following a logistic regression model including all
baseline covariates, i.e.  $P(Y(0)=1\given X=x) = \text{expit}(lp_0) =
e^{lp_0}/(1+e^{lp_0})$, with $lp_0=lp_0(x)=x^t\beta$.  In the base scenarios
coefficient values $\beta$ were such, that the ~~AUC of the logistic regression
model was 0.75 and the event rate in the control arm was
$20\%$.~~

:::{custom-style="bluefont"}
the control event rate was 20\% and the discriminative
ability of the true prediction model measured using Harrell's c-statistic was
0.75. The c-statistic represents the probability that for a randomly selected
discordant pair from the sample (patients with different outcomes) the
prediction model assigns larger risk to the patient with the worse outcome. For
the simulations this was achieved by selecting $\beta$ values such that the true
prediction model would achieve a c-statistic of 0.75 in a simulated control arm
with 1,000,000 patients. In the base case scenario we achieved that by setting
$\beta=(-2.08, 0.49,\dots,0.49)^t$.
:::

Outcomes in the treatment arm were first generated using 3 simple scenarios:
absent (OR = 1), moderate (OR = 0.8) or strong (OR = 0.5) constant relative
treatment effect. We then introduced linear, quadratic and non-monotonic
deviations from constant treatment effects using: $$lp_1 = \gamma_0 +
\gamma_1(lp_0-c) + \gamma_2(lp_0-c)^2, $$ where $lp_1$ is the true linear
predictor in the treatment arm, so that $P(Y(1)=1\given X=x) =
\text{expit}(lp_1)$,

:::{custom-style="bluefont"}
$\gamma=(\gamma_0, \gamma_1, \gamma_2)^t$
controls the shape of evolution of treatment effect as a function of baseline
risk (type and strength of deviations from the constant treatment effect
setting), while $c$ allows us to shift the proposed shape function to achieve
the desired overall event rates. For example, to simulate a constant treatment
effect with $\text{OR}=0.8$ we would set $\gamma=(\log(0.8), 1, 0)^t$ and
$c=0$.
:::

Finally, we incorporated constant absolute harms for all treated
patients, such that $P(Y(1)=1|X=x) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 (80\% power for the
detection of a marginal OR of 0.8 ~~with the standard alpha of 5\%~~

:::{custom-style="bluefont"}
using a 5\% significance level for the Wald test for the treatment coefficient
in a logistic regression model).
:::

We evaluated the
~~effect~~

:::{custom-style="bluefont"}
impact
:::

of smaller or larger sample sizes of 1,063
and 17,000, respectively. We also evaluated the ~~effect~~

:::{custom-style="bluefont"}
impact
:::

of risk model discriminative ability, adjusting the
baseline covariate coefficients, such that the AUC of the regression model in
the control arm was 0.65 and 0.85, respectively.

@@

Second, we ~~considered~~

:::{custom-style="bluefont"}
fitted
:::

a

:::{custom-style="bluefont"}
logistic regression 
:::

model which assumes *constant relative treatment effect* (constant
odds ratio)

:::{custom-style="bluefont"}
, that is, $P(Y=1\given X=x, Z=z;\hat{\beta}) =
\text{expit}(\hat{lp}_0 + \delta_0z)$.
:::

Hence, absolute benefit is predicted
from $\tau(x;\hat{\beta}) = \text{expit}(\hat{lp}_0) -
\text{expit}(\hat{lp}_0+\delta_0)$, where $\delta_0$ is the log of the assumed
constant odds ratio and $\hat{lp}_0 = \hat{lp}_0(x;\hat{\beta}) =
x^t\hat{\beta}$ the linear predictor of the estimated baseline risk model.

Third, we ~~considered~~

:::{custom-style="bluefont"}
fitted
:::

a logistic regression model
including treatment, the prognostic index, and their linear interaction

:::{custom-style="bluefont"}
, that is, $P(Y=1\given X=x, Z=z;\hat{\beta}) =
\text{expit}(\delta_0+\delta_1z+\delta_2
\hat{lp}_0+\delta_3z\hat{lp}_0)$
:::

Absolute benefit is then estimated from
$\tau(x;\hat{\beta}) = \text{expit}(\delta_0+\delta_1\hat{lp}_0) -
\text{expit}(\delta_0 + \delta_1\hat{lp}_0 + \delta_2z+\delta_3z\hat{lp}_0)$. We
will refer to this method as the *linear interaction* approach.

@@

We compared the discriminative ability of the methods under study using
c-for-benefit and the integrated calibration index (ICI) for
benefit ~~(Supplement, section 6)~~.

:::{custom-style="bluefont"}
The c-for-benefit represents the probability that from two randomly chosen
matched patient pairs with unequal observed benefit, the pair with greater
observed benefit also has a higher predicted benefit. To be able to calculate
observed benefit, patients in each treatment arm are ranked based on their
predicted benefit and then matched 1:1 across treatment arms. Observed treatment
benefit is defined as the difference of observed outcomes between the untreated
and the treated patient of each matched patient pair. Predicted benefit is
defined as the average of predicted benefit within each matched patient pair
[@vanKlaveren2018].
:::

:::{custom-style="bluefont"}
We evaluated calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. The observed benefits are regressed on
the predicted benefits using a locally weighted scatterplot smoother
(loess). The ICI-for-benefit is the average absolute difference between
predicted and smooth observed benefit. Values closer to represent better
calibration.
:::

@@

We demonstrated the different methods using `r nrow(gusto)` patients with acute
myocardial infarction (MI) included in the GUSTO-I trial.  `r treatmentArms %>%
filter(treatment == 1) %>% pull(n)` patients were randomized to tissue
plasminogen activator (tPA) treatment and `r treatmentArms %>% filter(treatment
== 0) %>% pull(n)` were randomized to streptokinase. The outcome of interest was
30-day mortality (total of `r sum(gusto$outcome)` events), recorded for all
patients.

:::{custom-style="bluefont"}
This dataset has been used extensively in prior studies
:::

~~In line with previous analyses~~ [@Califf1997; @Steyerberg2000]

:::{custom-style="bluefont"}
Therefore, we used the same set of 7 covariates that was previously used to,
:::

~~we fitted~~

:::{custom-style="bluefont"}
fit
:::

a logistic regression
model with ~~6~~

:::{custom-style="bluefont"}
the following
:::

baseline covariates: age, Killip class, systolic blood pressure, heart rate, an
indicator of previous MI, and the location of MI, to predict 30-day mortality
risk (Supplement, Section 8).

@@

The constant treatment effect approach outperformed other approaches in the base
case scenario (N = `r 4250`; OR = 0.8; AUC= 0.75; no absolute treatment harm)
with a true constant treatment effect (median RMSE: constant treatment effect 
`r median(rmseBaseCase$constant_treatment_effect)`;
linear interaction `r median(rmseBaseCase$linear_predictor)`;
RCS-3 `r median(rmseBaseCase$rcs_3_knots)`). 
The linear interaction model was optimal under true linear deviations 
(median RMSE: constant treatment effect 
`r median(rmseLinearInteraction$constant_treatment_effect)`;
linear interaction `r median(rmseLinearInteraction$linear_predictor)`;
RCS-3 `r median(rmseLinearInteraction$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
and even in the presence of true quadratic deviations
(median RMSE: constant treatment effect 
`r median(rmseQuadratic$constant_treatment_effect)`;
linear interaction `r median(rmseQuadratic$linear_predictor)`;
RCS-3 `r median(rmseQuadratic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
from a constant relative treatment effect. With non-monotonic deviations, RCS-3
slightly outperformed the linear interaction model
(median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
With strong treatment-related harms the results were very similar in most
scenarios (Figure \ref{fig:rmsebase} panels A-C). Under non-monotonic deviations
the optimal performance of RCS-3 was more pronounced
(median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
A stronger average treatment effect (OR=0.5)

:::{custom-style="bluefont"}
resulted in higher variability of the true treatment effects on the absolute
scale (difference in true outcome probabilities between treatment arms)
:::

~~led to larger absolute benefit
predictions~~ and consequently to larger RMSE for all approaches. ~~but~~

:::{custom-style="bluefont"}
However,
:::

the relative differences between different approaches were similar to the base case scenario
(Supplement, Figure S10). 
```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot on the top. The 2.5, 25, 50, 75, and 97.5 percentiles of the true benefit distributions are expressed by the boxplots on the side of the right-handside panel.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

@@

```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250 using the c-statistic for benefit. The c-statistic for benefit represents the probability that from two randomly chosen matched patient pairs with unequal observed benefit, the pair with greater observed benefit also has a higher predicted benefit. True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

@@

:::{custom-style="bluefont"}
Predicted absolute benefit was derived as the difference of predicted acute MI
risk between treatment arms, if all other predictors remained unchanged.
:::

All considered methods provided similar fits, predicting increasing

:::{custom-style="bluefont"}
absolute
:::

benefits for patients with higher baseline risk
predictions, and followed the evolution of the stratified estimates closely
(Figure \ref{fig:gusto}). The constant treatment effect model had somewhat lower
AIC compared to the linear interaction model (AIC:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(aic) %>% round()` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(aic) %>% round()`),
equal cross-validated discrimination
(c-for-benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(discrimination)`),
and slightly better cross-validated calibration
(ICI-for benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(calibration)` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(calibration)`).
In conclusion, although the sample size (`r nrow(gusto)` patients; 
`r sum(gusto$outcome)` events) allowed for flexible modeling approaches, a 
simpler constant treatment effect model is adequate for predicting absolute
30-day mortality benefits of treatment with tPA in patients with acute MI.
